ExpressionView - an interactive viewer for biclusters identified in gene expression data

### 2009-09-22 expressionviewnote/expressionview.tex

\section{Summary:}
ExpressionView is a Bioconductor package that provides an interactive environment to explore biclusters identified in gene expression data. A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results. From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it. The package is aimed to facilitate the exchange between bioinformaticians and life scientists who are not familiar with Bioconductor and the R language.

\section{Availability:}
\href{http://www.bioconductor.org/packages/2.0/bioc/html/ExpressionView.htmI}{Bioconductor package}

@@

\section{Introduction}
Clustering genes according to their expression profiles is an important task in analysing microarray data. While there is a multitude of biclustering software available~\citep{?}, none of them provides an intuitive interface that would allow the user to interactively explore the obtained results. ExpressionView is designed to close this gap, hereby facilitating the exchange between bioinformaticians and life scientists without consolidated programming experience. Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Flash applet that allows him and his co-workers to  visualise the biclusters together with the underlying gene expression data.

\section{Package design and workflow}
The ExpressionView package contains two independent parts - an ordering algorithm implemented in C++ and a Flash applet written in ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with biclustering results to produce an ExpressionView data file that can be explored with the Flash applet. In the following, we describe the different steps in more detail.
\begin{figure*}[!tpb]
\centerline{\includegraphics[width=\linewidth]{fig1-crop}}
\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples. In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results. The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
\end{figure*}
For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://maya:7575/ExpressionView}{website}.

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

\subsection{Order}
To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to regroup the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles. Since for more than two mutually overlapping biclusters, it is in general impossible to find such an arrangement, one has to make concessions. In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the area of the largest contiguous biclusters.

Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here. Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters. The quality of the order, which we call fragmentation, is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements. Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.

To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases. We find that in  almost every situation, the proposed algorithm finds a solution that recovers more than 99\% of the optimal fragmentation and in most cases, it actually recovers the correct alignment. In these test samples, execution time increases as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, depending on the number of elements $n$.

\subsection{Export}
Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into a single uncompressed XML file. This format has the advantage of being self-explanatory and extensible. In addition, the files can be read and edited with any text editor. The actual gene expression data is rounded to two digit precision and encoded in Base64. The structure of the XML file is described in more detail on the ExpressionView \href{http://maya:7575/ExpressionView}{website}. In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.

\subsection{Visualisation}
The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data, the biclusters, and the results of the enrichment calculations. Relying on the flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid. On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations are shown. Wherever possible, these elements are linked to the corresponding databases. The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item. The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\

@@

\bibitem[Madeira and Oliveira(2004)]{madeira04}
Madeira,S.C. and Oliveira,A.L. (2004)
Biclustering algorithms for biological data analysis: a survey.
\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.

>>>>

\section{Summary:}
ExpressionView is an R package that provides an interactive environment to explore biclusters identified in gene expression data. A sophisticated ordering algorithm is used to present the biclusters in a visually appealing layout that provides an intuitive summary of the results. From this overview, the user can select individual biclusters and access all the biologically relevant data associated with it. The package is aimed to facilitate the collaboration between bioinformaticians and life scientists who are not familiar with the R language.

\section{Availability:} 
\href{http://www.google.com}{R package with open source license}

@@

\section{Introduction}
Biclustering is an unsupervised learning technique used to analyse microarray data. It consists in identifying co-regulated groups of genes and conditions according to their expression profile. While there is a multitude of biclustering software available~\citep{madeira04}, packages with intuitive interfaces that allow the user to interactively explore the results are sparse~\citep{santamaria08}. ExpressionView is designed to close this gap, hereby facilitating the collaboration between bioinformaticians and life scientists without consolidated programming experience. Implemented as an R package, the user can apply all the powerful microarray analysis tools provided by the Bioconductor~\citep{gentleman04} project before exporting the data to a platform independent Adobe Flash applet that allows him and his co-workers to visualise the biclusters together with the underlying gene expression data.

\section{Package design and workflow}
The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and a Adobe Flash applet written in ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet. In the following, we describe the different steps in more detail.
\begin{figure*}[!tpb]
\centerline{\includegraphics[width=\linewidth]{fig1-crop}}
\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples. In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results. The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
\end{figure*}
For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://maya:7575/ExpressionView}{website}.

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

\subsection{Order}
To present the tens of possibly overlapping biclusters in a visually appealing form, it is necessary to reorder the rows and columns of the gene expression matrix in such a way that biclusters form contiguous rectangles. Since for more than two mutually overlapping biclusters, it is in general impossible to find such an arrangement, one has to make concessions. In contrast to \cite{grothaus06}, who proposed to repeat rows and columns as necessary to achieve this goal, we prefer to optimise the arrangement within the original data by maximising the area of the largest contiguous biclusters.

Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here. Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters. The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements. Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.

To determine the efficiency of this method, we have studied a large number of perfectly orderable, but initially scrambled, test cases. We find that in almost every situation, the proposed algorithm finds an order that recovers more than 99\% of the score of the optimal solution and in most cases, it actually recovers the correct alignment. In these test samples, execution time increases polynomially with the number of clusters $m$ as ${\mathcal O}(m^\alpha)$, where $\alpha \in [4, 4.5]$, almost independent of the number of elements $n$.

\subsection{Export}
Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into an XML file. This format has the advantage of being self-explanatory and extensible. In addition, the files can be read and edited with any text editor. The actual gene expression data is rounded to two digit precision and encoded in Base64. The structure of the XML file is described in more detail on the ExpressionView \href{http://maya:7575/ExpressionView}{website}. In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.

\subsection{Visualisation}
The core application of this package is the Flash applet that allows the user to interactively explore the gene expression data and the biclusters. Relying on the Flash architecture has the advantage of producing a platform and browser independent applet that can be launched from any computer with even the most restrictive user rights. A screenshot is shown in Fig.~\ref{fig:workflow}. The interface is divided in two parts: On the left-hand side, the user finds the gene expression data in the common heat map form, on top of which the biclusters are overlaid. On the right-hand side, the metadata associated with the gene expression data and the results of the enrichment calculations (GO and KEGG) are shown. Wherever possible, these elements are linked to the corresponding databases. The interface essentially behaves as an image viewer, allowing the user to zoom and pan around the expression data, getting instant feedback on the selected item. The currently selected region can be exported to a PDF file, helpful to prepare figures for presentations or articles, or a simple text file.\\

@@

\bibitem[Madeira and Oliveira(2004)]{madeira04}
Madeira,S.C. and Oliveira,A.L. (2004)
Biclustering algorithms for biological data analysis: A survey.
\href{http://dx.doi.org/10.1109/TCBB.2004.2}{\emph{IEEE/ACM Transactions on Computational Biology and Bioinformatics}, {\bf 1}, 24 -- 45.}

\bibitem[Prelic et~al.(2006)]{prelic06}
Prelic,A., \emph{et~al.} 
A systematic comparison and evaluation of biclustering methods for gene expression data 
\href{http://dx.doi.org/10.1093/bioinformatics/bt1060}{\emph{Bioinformatics}, {\bf 22}(9), 1122 -- 1129.}

\bibitem[Santamar\'ia et~al.(2008)]{santamaria08}
Santamar\'ia,R., Ther\'on,R. and Quintales,L. (2008)
BicOverlapper: A tool for bicluster visualization. 
\href{http://dx.doi.org/10.1093/bioinformatics/btn076}{\emph{Bioinformatics}, {\bf 24}, 1212 -- 1213.}
